---
title: "RepData_PA1_WJ"
output: html_document
---

This is my R Markdown document for Peer Assessment 1 of the Reproducible Research Course provided by JHU

### Setting Global Options
```{r setoptions, echo=TRUE}
library(knitr)
opts_chunk$set(echo=TRUE)
``` 



###Loading Data
Here, we're going to load the activity monitoring data.
```{r loaddata}
unzip("./activity.zip",exdir = ".")
activity<-read.csv("./activity.csv")
```



### Mean Total Number of Steps Taken per Day
First, get rid of the missing value in the dataset
```{r }
activity_new<-activity[complete.cases(activity),]
```

Secondly, summarize the total steps taken on each day and make a histogram
```{r }
daily_step<-aggregate( steps ~ date, data=activity_new, FUN=sum)
hist(daily_step$steps, col="blue",breaks=10, main="Steps Taken Per Day", xlab="Steps")
```

Then, calculate the mean and median of the total steps
```{r }
mean_steps<-mean(daily_step$steps)
median_steps<-median(daily_step$steps)
```
The mean of total steps taken per day is `r mean_steps`.  The median of total steps taken per day is `r median_steps`



### Average Daily Activity Pattern and Make a Histogram
Average the steps based on intervals, i.e steps taken in each five minutes
```{r }
interval_step<-aggregate( steps ~ interval, data=activity_new, FUN="mean")
plot(interval_step$interval, interval_step$steps,col="blue", main="Steps Taken in Each Five Minutes", 
     xlab="Five Minutes Interval",ylab="Avg Steps in Each Five Minutes", type="l")
```

Find the interval with maximum steps across all the  days
```{r}
maximum_step<-max(interval_step$steps)
for (i in 1:nrow(interval_step)){
  if (interval_step[i,2]==maximum_step){
    maximum_interval<-interval_step[i,1]
  }
}
```
The 5-minute interval with maximum steps is `r maximum_interval`



### Imputing Missing Values
Calculate the total number of missing values in the dataset
```{r}
missing_number<-nrow(activity)-nrow(activity_new)
```
The total number of rows with NAs is `r missing_number`

Create a new dataset with filled missing values. 
Here, I use the mean of the steps taken in 5-minutes intervals as the replacement for the missing data
```{r}
activity_fill<-activity
library(hash)
h<-hash(keys=interval_step$interval,values=interval_step$steps)
for (i in 1:nrow(activity_fill)){
  
  if (is.na(activity_fill[i,1])){
    hash_key<-as.character(activity_fill[i,3])
    activity_fill[i,1]<-as.numeric(h[[hash_key]])
    
  }
}
```

Then, make a histogram of the total number of steps taken each day with filled values
```{r}
daily_step_fill<-aggregate( steps ~ date, data=activity_fill, FUN=sum)
hist(daily_step_fill$steps, col="blue",breaks=10, main="Steps Taken Per Day", xlab="Steps")
```

Calculate the new mean and new median of the total steps
```{r fig.width=10 }
mean_steps_filled<-mean(daily_step_fill$steps)
median_steps_filled<-median(daily_step_fill$steps)
par(mfrow=c(1,2))
hist(daily_step$steps, col="blue",breaks=10, main="Steps Taken Per Day-Before Imputation", xlab="Steps")
hist(daily_step_fill$steps, col="blue",breaks=10, main="Steps Taken Per Day-After Imputation", xlab="Steps")
```

The new mean of total steps taken per day is `r mean_steps_filled`.  The new median of total steps taken per day is `r median_steps_filled`. Compare the histogram side by side, it seems that the imputation doesn't impact the mean and median values much (only one step increase in median value) and the general distribution shape has been preserved. However, after the imputation, the missing dates have been filled averaged value and the variance of the total daily steps has been reduced. 




### Activity Patterns in Weekdays and Weekends

First, create a new factor variable in the dataset indicating whether a given date is a weekday or weekend day
```{r}
activity_fill$date<-as.Date(activity_fill$date)

for(i in 1:nrow(activity_fill)){
  if((weekdays(activity_fill[i,2])=="Saturday")||(weekdays(activity_fill[i,2])=="Sunday")){
    activity_fill[i,4]="Weekend"
  } else {
    activity_fill[i,4]="Weekday"    
  }
}
colnames(activity_fill)[4]<-"week"
activity_fill$week<-as.factor(activity_fill$week)
head(activity_fill,3)
```

Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
```{r}
interval_step_week<-aggregate( steps ~ interval+week, data=activity_fill, FUN="mean")
library(ggplot2)
g<-ggplot(interval_step_week,aes(interval,steps))
g+geom_line(color="steelblue",size=0.7)+facet_grid(.~week)
```

Based on the panel plot, it can be inferred that this person has more activities in weekdays than that in weekends